Appendix: Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-02-27 Appendix_plain_v3.tex
Updates from Overleaf

\begin{align} 
	 log(F_s(x,y,d)) = \kappa + \eta(x,y) + \psi_{d}(x,y,d) \\
	 G_{s}(x,y,d) \sim \mathrm{Tweedie}(F_s(x,y,d),p ,\theta_s)
\end{align}
where x, y, and d index longitude, latitude, and depth, respectively.   $\kappa$ is scalar intercept. Both $\eta(x,y)$ and $\psi_d(x,y,d)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim  \mathrm{MVN}(0,\sum_\eta)$ and $\psi_d \sim  \mathrm{MVN}(0,\sum_{\psi_d})$ (see \cite{anderson2022}) where $\eta$ is a depth-invariant spatial field and $\psi_d$ is a spatial field specific to each depth category. In simple terms, each location-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated DNA, while $\psi_d$ modulates that estimate for different sampling depths. The $\psi_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$.

Estimation of large, multi-level spatiotemporal models can strain computational power, in part because of the need in spatial models to invert large covariance matrices between the data. We use the Stochastic Partial Differential Equation (SPDE) model, implemented through the fmesher R package (\cite{lindgren2024}; 10.32614/CRAN.package.fmesher) to approximate the latent spatial process and its variance by using a triangulated mesh (\cite{lindgren2011}). The SPDE approach drastically reduces computation time by using a sparse covariance matrix to approximate Gaussian Markov random fields like $\eta$ and $\psi_d$. Model estimates and the degree of spatial smoothing can be affected by the details of mesh construction, and we therefore produced a range of meshes with different complexities (i.e., varying numbers of triangles, triangle sizes, and therefore vertices) to test for each species.

>>>>

\begin{align} 
	 log(F_s(x,y,d)) = \kappa + \eta(x,y) + \omega_{d}(x,y,d) \\
	 G_{s}(x,y,d) \sim \mathrm{Tweedie}(F_s(x,y,d),p ,\theta_s)
\end{align}
where x, y, and d index longitude, latitude, and depth, respectively.   $\kappa$ is scalar intercept. Both $\eta(x,y)$ and $\omega_d(x,y,d)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim  \mathrm{MVN}(0,\sum_\eta)$ and $\omega_d \sim  \mathrm{MVN}(0,\sum_{\omega_d})$ (see \cite{anderson2022}) where $\eta$ is a depth-invariant spatial field and $\omega_d$ is a spatial field specific to each depth category. In simple terms, each location-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated DNA, while $\omega_d$ modulates that estimate for different sampling depths. The $\omega_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$.

Estimation of large, multi-level spatiotemporal models can strain computational power, in part because of the need in spatial models to invert large covariance matrices between the data. We use the Stochastic Partial Differential Equation (SPDE) model, implemented through the fmesher R package (\cite{lindgren2024}; 10.32614/CRAN.package.fmesher) to approximate the latent spatial process and its variance by using a triangulated mesh (\cite{lindgren2011}). The SPDE approach drastically reduces computation time by using a sparse covariance matrix to approximate Gaussian Markov random fields like $\eta$ and $\omega_d$. Model estimates and the degree of spatial smoothing can be affected by the details of mesh construction, and we therefore produced a range of meshes with different complexities (i.e., varying numbers of triangles, triangle sizes, and therefore vertices) to test for each species.
